---
title: "MEDBIOINFO: Statistics for Bioinformatics -- CCA examples"
author: "Dirk.Repsilber@oru.se"
date: '2025-11-27'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(PMA)
```

# Data from APRO study, log-transformed and imputed

Data are from the APRO study, to be published, from the NGBI Research Environment at the School of Medical Sciences, Örebro University:

https://www.oru.se/english/research/research-environments/mh/nutrition-gut-brain-interactions-research-centre-ngbi/Research_Teams/

X: questionnaire data ("symptoms")

Z: blood markers for inflammation, gut permeability, tissue damage, oxidative stress, SCFAs


```{r}
X <- readRDS("APRO_X.RDS")
Z <- readRDS("APRO_Z.RDS")
```


# sCCA with X markers, Z outcome

```{r}
## number of latent variable pairs to extract:
my.K <- 10 
standardize <- FALSE

CCA.res <- CCA(x=X ,z=Z, standardize=standardize, K=my.K)

## canonical correlations found:
CCA.res$cors 

## reproduce 5th pair of latent variables and its correlation:
cor(X %*% CCA.res$u[,5], Z %*% CCA.res$v[,5])
```

# Permutation p-values

Function to implement permutation test for canonical correlations:

```{r}
get.perm.p.CCA <- function(my.x,my.z,K,N,standardize,seed){
  require(PMA)
  CCA.res.orig <- CCA(x=my.x,z=my.z,standardize=standardize,K=K,trace=FALSE)
  CCA.cor.orig <- CCA.res.orig$cors
  CCA.perm.cors <- matrix(NA,nrow=N,ncol=K)
  set.seed(seed)
  for(p in 1:N){
    my.x.perm <- my.x[sample(1:nrow(my.x)),]
    my.z.perm <- my.z[sample(1:nrow(my.z)),]
    CCA.perm.cors[p,] <- (CCA(x=my.x.perm,z=my.z.perm,standardize=standardize,K=K,trace=FALSE))$cors
  }
  p.values <- c()
  for(k in 1:K){
    p.values[k] <- sum(CCA.perm.cors[,k] >= CCA.cor.orig[k])/N
  }
  return(list(CCA.cor=CCA.cor.orig,p=p.values))
}
```

```{r}
res.perm   <- get.perm.p.CCA(my.x=X,my.z=Z,K=my.K,N=500,standardize=standardize,seed=123)

res.perm
```


# Scatterplot of latent variables

```{r}
no.CCA <- 5

plot(X %*% CCA.res$u[,no.CCA], Z %*% CCA.res$v[,no.CCA], main=paste("CCA cor=",round(CCA.res$cors[no.CCA],digits=3)))
abline(lm(Z %*% CCA.res$v[,no.CCA] ~ X %*% CCA.res$u[,no.CCA]))

```

# Extracting important variables

```{r}
for(i in 1:my.K){
  print("-----------------")
  print(paste("latent pair: ",i))
  print(paste("u variables:"))
  print(colnames(X)[which(CCA.res$u[,i] != 0)])
  print(paste("v variables:"))
  print(colnames(Z)[which(CCA.res$v[,i] != 0)])
  print("-----------------")
}

```









